---
title: "Group Project 4"
subtitle: "Biology 368/664 Bucknell University"
output: html_notebook
author: Prof. Ken Field
date: 9 Feb 2022
---

```{r Load Libraries, include=FALSE}
# Load other packages here.
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse)
```

## Target Audience

Discuss with your group the target audience for the tutorial. 
Examples could be one of the new core Biology classes, another 300-level course (not 364), or a research group. 

```{r echo=FALSE}

```

## Grading

Each student will be expected to complete the following tasks to earn 85% of the points available for this assignment (21/25).

- Identify and obtain suitable dataset
- Use a Github repository and version control to collaborate on the project
- Spend 4-6 hours preparing, coding, and testing tutorial
  + Data exploration: Sam
  + Data visualization: Irene
  + Hypothesis testing: Abby 
- Present tutorial in class
- Provide public archive suitable for sharing to students/faculty

Tutorials from previous classes can be viewed at our public github site: https://github.com/Bucknell-Biol364

Each group should use an *Acknowledgements* section to document the participation of each member and the collaboration within and between groups.

Additional credit will be awarded for providing assistance to other groups and for the development of a tutorial that goes beyond the minimal expectations listed above.

#Introducing R

#Data exploration 
The first thing to do when exploring data is to look at the whole datasheet. This can be done with the summary command. 

*summary() allows you to look at the summary of the data*
```{r}
summary(msleep)
```

You can also view the data in table form to see if it imported correctly 

*Just type the name of the data set*
```{r}
msleep
```

Now that the whole data set has been seen you can now begin to look at different variables.  

For the hypothesis we are looking at ____ so we can look at a box plot of the data. Type the command below.

*ggplot(msleep) + aes(x = ) + geom_histogram() + theme_cowplot()*
```{r}
ggplot(msleep) +
  aes(x = brainwt) +
  geom_histogram() +
  theme_cowplot()
```

From this box plot we can see the normality of the data. It is important that the data be distributed normally because for statistical tests you will see later many assume the data is normal.

We can see here that the data is not distrusted normally, so we can try and transform the to become normally distrusted. 

Try making a histogram of ____ but this time try doing a log transformation. 

*Type the same command as before but this time for x type log(____)*
```{r}
ggplot(msleep) +
  aes(x = log(brainwt)) +
  geom_histogram() +
  theme_cowplot()
```

The data was not made normal from a transformation. This happens sometimes it is okay. Just make a note when doing a statistical test that the data was not distributed normally.

If you had a more specific hypothesis and wanted to look at a more specific range within the variable you could do that. Maybe you wanted to only look at the brain weights were the body weight was below 100.

*new name <- filter(msleep, (bodywt <= 100))*
*ggplot(new name) + aes(x = brainwt) + geom_histogram() + theme_cowplot()*
```{r}
bodywt.large <- filter(msleep, (bodywt <= 100))

ggplot(bodywt.large) +
  aes(x = brainwt) +
  geom_histogram() +
  theme_cowplot()
```

If you needed, you could do multiple data filters for different variable at the same time. Just make sure to give the filtered data good name. Sometimes you can end up with a lot of different names, so stay organized. Also you never want to name something the same as the name of the data set because R will overwrite the data with your filtered data. 

When you explore the data enough you can move onto truly visualizing it. 

#Data visulization 

#Hypothesis Testing 


## Sample Dataset
One of the possible datasets to use for the tutorial can be found in the datasauRus package.

```{r}
datasaurus_dozen %>% 
  group_by(dataset) %>% 
  summarize(
      mean_x    = mean(x),
      mean_y    = mean(y),
      std_dev_x = sd(x),
      std_dev_y = sd(y),
      corr_x_y  = cor(x, y)
      )
```

Boxplots of either the x or the y value show that there is something strange going on.

```{r}
ggplot(datasaurus_dozen, aes(x = x, colour = dataset))+
    geom_boxplot()+
    theme_void()+
    theme(legend.position = "none")+
    facet_wrap(~dataset, ncol = 3)
```

```{r}
ggplot(datasaurus_dozen, aes(x = y, colour = dataset))+
    geom_boxplot()+
    theme_void()+
    theme(legend.position = "none")+
    facet_wrap(~dataset, ncol = 3)
```

But you have to visualize all of the data with a scatter plot to really see the patterns.

```{r fig.height=12, fig.width=9}
  ggplot(datasaurus_dozen, aes(x = x, y = y, colour = dataset))+
    geom_point()+
    theme_void()+
    theme(legend.position = "none")+
    facet_wrap(~dataset, ncol = 3)
```

And did you notice the code in the {r} codechunk header that controlled the size of the output in the Rmd? Pretty neat trick!

And here are two versions of the data that you could use in your data visualization tutorial. 
To use them you would probably want to change the names of the datasets and also make x and y more meaningful. 
Then save them as a csv or tsv to be imported later for your tutorial. 

```{r}
datasaurus_long <- datasaurus_dozen
datasaurus_wide <- datasaurus_dozen_wide
head(datasaurus_long)
head(datasaurus_wide)
```

# Acknowledgements

DatasauRus package and description below: Stephanie Locke https://github.com/jumpingrivers/datasauRus

The datasauRus package wraps the awesome Datasaurus Dozen dataset, which contains 13 sets of x-y data. Each sub-dataset has five statistics that are (almost) the same in each case. (These are the mean of x, mean of y, standard deviation of x, standard deviation of y, and Pearson correlation between x and y). However, scatter plots reveal that each sub-dataset looks very different. The dataset is intended to be used to teach students that it is important to plot their own datasets, rather than relying only on statistics.

The Datasaurus was created by Alberto Cairo in this great [blog post](http://www.thefunctionalart.com/2016/08/download-datasaurus-never-trust-summary.html).

He's been subsequently made even more famous in the paper [Same Stats, Different Graphs: Generating Datasets with Varied Appearance and Identical Statistics through Simulated Annealing](https://www.autodeskresearch.com/publications/samestats) by Justin Matejka and George Fitzmaurice.

In the paper, Justin and George simulate a variety of datasets that the same summary statistics to the Datasaurus but have very different distributions. 

This package also makes these datasets available for use as an advanced [Anscombe's Quartet](https://en.wikipedia.org/wiki/Anscombe%27s_quartet), available in R as `anscombe`.


Team Members: 
Irene Hu 


